Bucket-Sorted Independent Sets for Algebraic Multigrid

### 2007-08-06 bsis_paper/paper.tex
* Added image files.
git-svn-id: file:///home/rblake/svn/pamg/trunk@13 c7f3fe13-8c30-0410-9127-9a9c42fe009c
* This document class is strange and seems to be causing various   problems. Not sure what caused Luke to use this style. Need to   ask. * Cleaned up some of the problems with the references. * Cleaned up title page. * Thought about the design of the document and put some of that in   the TODO file. Began making changes in document structure in the   .tex file.
git-svn-id: file:///home/rblake/svn/pamg/trunk@14 c7f3fe13-8c30-0410-9127-9a9c42fe009c

% if you use PostScript figures in your article
% use the graphics package for simple commands
% \usepackage{graphics}
% or use the graphicx package for more complicated commands
\usepackage[draft]{graphicx}
% or use the epsfig package if you prefer to use the old commands
% \usepackage{epsfig}

@@

% use the thanksref command within \title, \author or \address for footnotes;
% use the corauthref command within \author for corresponding author footnotes;
% use the ead command for the email address,
% and the form \ead[url] for the home page:
\title{Bucket-Sorted Independent Sets for Algebraic Multigrid}
\author{David Alber\corauthref{cor}},
\corauth[cor]{Corresponding author. Address: 201 N. Goodwin Ave., Siebel
Center for Computer Science, Urbana, IL 61801}
\ead{dalber@uiuc.edu}
\ead[url]{http://www.cs.uiuc.edu/homes/dalber}
\author{Luke Olson}
\address{Department of Computer Science\\ University of Illinois at Urbana-Champaign}
\ead{dalber@uiuc.edu}
\ead[url]{http://www.cs.uiuc.edu/homes/lukeo}
% \thanks[label2]{}
% \address{Address\thanksref{label3}}
% \thanks[label3]{}
% use optional labels to link authors explicitly to addresses:
% \author[label1,label2]{}
% \address[label2]{}

@@

@@

In each iteration, the non-empty bucket with largest weight forms $D$
(see Corollary~\ref{5:cor:max-weight-indep-set}). These vertices are
assigned to the $C$-point set and removed from the data
structure. Vertex weight updates lead to corresponding updates to the
data structure, and new $F$-points are removed from the data
structure. These operations continue until all buckets are empty, at
which point the coarse-grid selection is
complete. Algorithms~\ref{5:alg:BSIS-setup},
\ref{5:alg:BSIS-indep-set-select}, and~\ref{5:alg:BSIS-weight-update}
outline the operations discussed above.
\begin{lstlisting}[caption={BSIS Data Structure Setup},label=5:alg:BSIS-setup]
%\textsc{BSIS-Setup}$(S)$%
for $i \in V$
  $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
  $bucket[bucketID].\textsc{Insert}(i)$
end
\end{lstlisting}
% \begin{algorithm}[!h]
% \caption{BSIS Data Structure Setup}
% \label{5:alg:BSIS-setup}
% \textsc{BSIS-Setup}$(S)$ \{
% \begin{algorithmic}[1]
% \FORALL{$i \in V$}
% \STATE $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
% \STATE $bucket[bucketID].\textsc{Insert}(i)$
% \ENDFOR
% \end{algorithmic}
% \}
% \end{algorithm}
\begin{lstlisting}[caption={Independent Set Selection},label=5:alg:BSIS-indep-set-select]
%\textsc{BSIS-Independent-Set-Selection}$(S)$%
return non-empty $bucket$ with largest $bucketID$
\end{lstlisting}
% \begin{algorithm}[!h]
% \caption{Independent Set Selection}
% \label{5:alg:BSIS-indep-set-select}
% \textsc{BSIS-Independent-Set-Selection}$(S)$ \{
% \begin{algorithmic}[1]
% \RETURN non-empty $bucket$ with largest $bucketID$
% \end{algorithmic}
% \}
% \end{algorithm}
\begin{lstlisting}[caption={BSIS Weight Update},label=5:alg:BSIS-weight-update]
%\textsc{BSIS-Weight-Update}$(S)$%
$bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
$bucket[bucketID].\textsc{Remove}(i)$
$bucket[bucketID - numColors].\textsc{Insert}(i)$
\end{lstlisting}
%\begin{algorithm}[!h]
%\caption{BSIS Weight Update}
%\label{5:alg:BSIS-weight-update}
%\textsc{BSIS-Weight-Update}$(S)$ \{
%\begin{algorithmic}[1]
%\STATE $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
%\STATE $bucket[bucketID].\textsc{Remove}(i)$
%\STATE $bucket[bucketID - numColors].\textsc{Insert}(i)$
%\end{algorithmic}
%\}
%\end{algorithm}
\begin{figure}
  \begin{center}
    %\includegraphics[width=0.8\textwidth]{images/bucket-algorithm}
    \includegraphics[width=0.98\textwidth]{images/initial}
    \caption{The BSIS data structure.}
    \label{5:fig:bucket-algorithm}
  \end{center}
\end{figure}

Figure~\ref{5:fig:bucket-algorithm-cp1} illustrates the graph and data
structure following the first iteration of the algorithm. Vertex~10
has become a $C$-point and its neighbors weights have been
updated. Vertices assigned to $F$ or $C$ are removed from the data
structure and other affected vertices are moved to new locations in
the data structure. Such vertices are highlighted in red.
\begin{figure}
  \begin{center}
    \includegraphics[width=0.98\textwidth]{images/cp1}
    \caption{The BSIS data structure after selecting the first
    $C$-point (Vertex 10). The weights of neighbors of the new
    $C$-point are updated. Some neighbors become $F$-points and are
    removed from the data structure. Vertices removed from the data
    structure are highlighted with a red ring in the graph, while
    other neighbors are moved to new locations in the data structure
    and are highlighted (to help in reading the figure) in the data
    structure with a red box.}
    \label{5:fig:bucket-algorithm-cp1}
  \end{center}
\end{figure}

@@

Figure~\ref{5:fig:cp1-agg} depicts the data structure after the first
set of $C$-points is selected.
\begin{figure}
  \begin{center}
    \includegraphics[width=0.98\textwidth]{images/cp1-agg}
    \caption{The BSIS data structure after selecting the first
    $C$-point (Vertex 10) with aggregate weight updates.}
    \label{5:fig:cp1-agg}
  \end{center}
\end{figure}
Rather than moving vertices to new buckets, the method now keeps them
in the same location and only moves them when necessary. As shown in
Section~\ref{5:sec:experiments}, aggregation of the weight updates
leads to significant savings in computational cost.

@@

Timing data for the selection of all coarse grids in the hierarchy is
reported. This time includes the cost for all parts of the algorithms,
including the graph coloring phase. AMG solve phase information is not
reported since the algorithms produce identical coarse grids and since
information on solve phase performance for AMG with CLJP-c is
documented in~\cite{alber-cljpc, alber-PCGS}.

@@

The first idea is called the boundary painting method and takes
advantage of the invariance between coarse grids selected by BSIS and
CLJP-c. The idea is to use BSIS to select as many of the interior
$C$-points as possible before doing any communication with neighboring
processors. All vertices belonging to a processor are colored and
inserted into the BSIS data structure. The processor boundary
vertices, however, are ``painted''. If a painted vertex is in a set
$D$ in some iteration, then the vertex is not added to $C$. It is
instead removed from the data structure and its on-processor neighbors
are also painted. Figure~\ref{5:fig:cp1-parallel} illustrates the
first iteration of the painted boundary method with weight update
aggregation. The data structure shown is for the left domain.
\begin{figure}
  \begin{center}
    \includegraphics[width=0.98\textwidth]{images/cp1p}
    \caption{The painted boundary method with aggregate weight
    updates following one iteration. The data structure is for the
    left domain. Painted vertices are marked with a green ring in the
    graph and a green box in the data structure. Vertices in the data
    structure that are half green and half red are painted and also
    have had their weights updated.}
    \label{5:fig:cp1-parallel}
  \end{center}
\end{figure}
The first iteration selects a new $C$-point, but does not select any
painted vertices. In the second iteration, Vertex~22 is selected, but
is already painted. Therefore, on-processor neighbors of Vertex~22 are
also painted (see Figure~\ref{5:fig:cp2-parallel}).
\begin{figure}
  \begin{center}
    \includegraphics[width=0.98\textwidth]{images/cp2p}
    \caption{The painted boundary method with aggregate weight
    updates following the second iteration. In this iteration a
    painted vertex was selected, leading to the painting of its
    on-processor neighbors.}
    \label{5:fig:cp2-parallel}
  \end{center}
\end{figure}
The method finishes when the data structure is emptied of all
vertices. The product is now three disjoint sets: $C$ and $F$, as
usual, but also a set of painted vertices. The painted vertices are
the vertices that cannot be assigned to the $F$ or $C$ set without
communication with another processor. The information is provided to
CLJP-c, which handles the parallel portion of the algorithm.

The painted boundary approach is ideal given large numbers of interior
vertices, which can be guaranteed on most problems for the fine grid.
A side-effect of H1-based coarsening algorithms, however, is the
creation of denser graphs on coarse levels. One solution is to use the
painted boundary method on fine levels and then switch to CLJP-c
further along in the coarsening process. The issue is a smaller
concern for H1$^\prime$-based coarsening using BSIS since these
methods produce small operator complexities that do not grow as the
size of the problem is increased. In all cases, however, the number of
processor boundary vertices relative to the number of interior
vertices increases as the number of unknowns per processor decreases
(e.g., on coarse levels). A few techniques may be applicable in this
situation. The easiest solution is to simply use CLJP-c, or some other
coarsening algorithm, on the coarser levels where few vertices are
found. Although BSIS does not decrease in cost at this point, the
total cost on the levels when BSIS is not used is low due to low
complexity of coarse grids. A second approach is the application of a
dynamic load balancing algorithm to increase the number of vertices on
a processor (and, thus, decrease communication costs). If the number
of vertices per processor per level is maintained at a high enough
level, BSIS is still valuable on coarse grids. A third option is to
replicate the operator matrix on processors, which leads to processors
doing more of the same work, but by avoiding communication. The second
and third ideas are similar in nature and both involve using dynamic
load balancing techniques~\cite{DevineBomanKarypisPP04, CybDLB,
deCoughnyLB, SchloegelLB, ZoltanParHyp06ipdps, ZoltanParHypRepart07}.

>>>>

% if you use PostScript figures in your article
% use the graphics package for simple commands
% \usepackage{graphics}
% or use the graphicx package for more complicated commands
\usepackage{graphicx}
% or use the epsfig package if you prefer to use the old commands
% \usepackage{epsfig}

@@

% use the thanksref command within \title, \author or \address for footnotes;
% use the corauthref command within \author for corresponding author footnotes;
% use the ead command for the email address,
% and the form \ead[url] for the home page:
\title{Bucket-Sorted Independent Sets for Algebraic Multigrid}
\author[NREL]{David Alber\corauthref{cor}},
\corauth[cor]{Corresponding author. Address: 1617 Cole Blvd., Mail
  Stop: 1608, Golden, CO 80401}
\ead{david\_alber@nrel.gov}
%\ead[url]{http://www.cs.uiuc.edu/homes/dalber}
\author[UIUC]{Luke Olson}
\address[NREL]{Scientific Computing Center\\ National Renewable Energy Laboratory}
\address[UIUC]{Department of Computer Science\\ University of Illinois at Urbana-Champaign}
\ead{lukeo@uiuc.edu}
\ead[url]{http://www.cs.uiuc.edu/homes/lukeo}
% \thanks[label2]{}
% \address{Address\thanksref{label3}}
% \thanks[label3]{}
% use optional labels to link authors explicitly to addresses:
% \author[label1,label2]{}
% \address[label2]{}

@@

\section{Coarse Grid Selection}
\label{sec:coarse-grid-select}

\subsection{Problem}
\label{sec:problem}
% Outline coarse grid selection. Problem outline.

\subsection{Independent Set-Based Selection}
\label{sec:indep-set-based}
% Talk about independent set-based algorithms here, and give general
% independent set-based algrotihm here.

@@

In each iteration, the non-empty bucket with largest weight forms $D$
(see Corollary~\ref{5:cor:max-weight-indep-set}). These vertices are
assigned to the $C$-point set and removed from the data
structure. Vertex weight updates lead to corresponding updates to the
data structure, and new $F$-points are removed from the data
structure. These operations continue until all buckets are empty, at
which point the coarse-grid selection is
complete. Algorithms~\ref{5:alg:BSIS-setup},
\ref{5:alg:BSIS-indep-set-select}, and~\ref{5:alg:BSIS-weight-update}
outline the operations discussed above.
\begin{lstlisting}[caption={BSIS Data Structure Setup},label=5:alg:BSIS-setup]
%\textsc{BSIS-Setup}$(S)$%
for $i \in V$
  $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
  $bucket[bucketID].\textsc{Insert}(i)$
end
\end{lstlisting}
% \begin{algorithm}[!h]
% \caption{BSIS Data Structure Setup}
% \label{5:alg:BSIS-setup}
% \textsc{BSIS-Setup}$(S)$ \{
% \begin{algorithmic}[1]
% \FORALL{$i \in V$}
% \STATE $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
% \STATE $bucket[bucketID].\textsc{Insert}(i)$
% \ENDFOR
% \end{algorithmic}
% \}
% \end{algorithm}
\begin{lstlisting}[caption={Independent Set Selection},label=5:alg:BSIS-indep-set-select]
%\textsc{BSIS-Independent-Set-Selection}$(S)$%
return non-empty $bucket$ with largest $bucketID$
\end{lstlisting}
% \begin{algorithm}[!h]
% \caption{Independent Set Selection}
% \label{5:alg:BSIS-indep-set-select}
% \textsc{BSIS-Independent-Set-Selection}$(S)$ \{
% \begin{algorithmic}[1]
% \RETURN non-empty $bucket$ with largest $bucketID$
% \end{algorithmic}
% \}
% \end{algorithm}
\begin{lstlisting}[caption={BSIS Weight Update},label=5:alg:BSIS-weight-update]
%\textsc{BSIS-Weight-Update}$(S)$%
$bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
$bucket[bucketID].\textsc{Remove}(i)$
$bucket[bucketID - numColors].\textsc{Insert}(i)$
\end{lstlisting}
%\begin{algorithm}[!h]
%\caption{BSIS Weight Update}
%\label{5:alg:BSIS-weight-update}
%\textsc{BSIS-Weight-Update}$(S)$ \{
%\begin{algorithmic}[1]
%\STATE $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
%\STATE $bucket[bucketID].\textsc{Remove}(i)$
%\STATE $bucket[bucketID - numColors].\textsc{Insert}(i)$
%\end{algorithmic}
%\}
%\end{algorithm}
\begin{figure}
  \begin{center}
    %\includegraphics[width=0.8\textwidth]{images/bucket-algorithm}
    \includegraphics[width=0.98\textwidth]{images/BSIS/initial}
    \caption{The BSIS data structure.}
    \label{5:fig:bucket-algorithm}
  \end{center}
\end{figure}

Figure~\ref{5:fig:bucket-algorithm-cp1} illustrates the graph and data
structure following the first iteration of the algorithm. Vertex~10
has become a $C$-point and its neighbors weights have been
updated. Vertices assigned to $F$ or $C$ are removed from the data
structure and other affected vertices are moved to new locations in
the data structure. Such vertices are highlighted in red.
\begin{figure}
  \begin{center}
    \includegraphics[width=0.98\textwidth]{images/BSIS/cp1}
    \caption{The BSIS data structure after selecting the first
    $C$-point (Vertex 10). The weights of neighbors of the new
    $C$-point are updated. Some neighbors become $F$-points and are
    removed from the data structure. Vertices removed from the data
    structure are highlighted with a red ring in the graph, while
    other neighbors are moved to new locations in the data structure
    and are highlighted (to help in reading the figure) in the data
    structure with a red box.}
    \label{5:fig:bucket-algorithm-cp1}
  \end{center}
\end{figure}

@@

Figure~\ref{5:fig:cp1-agg} depicts the data structure after the first
set of $C$-points is selected.
\begin{figure}
  \begin{center}
    \includegraphics[width=0.98\textwidth]{images/BSIS/cp1-agg}
    \caption{The BSIS data structure after selecting the first
    $C$-point (Vertex 10) with aggregate weight updates.}
    \label{5:fig:cp1-agg}
  \end{center}
\end{figure}
Rather than moving vertices to new buckets, the method now keeps them
in the same location and only moves them when necessary. As shown in
Section~\ref{5:sec:experiments}, aggregation of the weight updates
leads to significant savings in computational cost.

@@

Timing data for the selection of all coarse grids in the hierarchy is
reported. This time includes the cost for all parts of the algorithms,
including the graph coloring phase. AMG solve phase information is not
reported since the algorithms produce identical coarse grids and since
information on solve phase performance for AMG with CLJP-c is
documented in~\cite{alber-cljpc,alber-PCGS}.

@@

The first idea is called the boundary painting method and takes
advantage of the invariance between coarse grids selected by BSIS and
CLJP-c. The idea is to use BSIS to select as many of the interior
$C$-points as possible before doing any communication with neighboring
processors. All vertices belonging to a processor are colored and
inserted into the BSIS data structure. The processor boundary
vertices, however, are ``painted''. If a painted vertex is in a set
$D$ in some iteration, then the vertex is not added to $C$. It is
instead removed from the data structure and its on-processor neighbors
are also painted. Figure~\ref{5:fig:cp1-parallel} illustrates the
first iteration of the painted boundary method with weight update
aggregation. The data structure shown is for the left domain.
\begin{figure}
  \begin{center}
    \includegraphics[width=0.98\textwidth]{images/BSIS/cp1p}
    \caption{The painted boundary method with aggregate weight
    updates following one iteration. The data structure is for the
    left domain. Painted vertices are marked with a green ring in the
    graph and a green box in the data structure. Vertices in the data
    structure that are half green and half red are painted and also
    have had their weights updated.}
    \label{5:fig:cp1-parallel}
  \end{center}
\end{figure}
The first iteration selects a new $C$-point, but does not select any
painted vertices. In the second iteration, Vertex~22 is selected, but
is already painted. Therefore, on-processor neighbors of Vertex~22 are
also painted (see Figure~\ref{5:fig:cp2-parallel}).
\begin{figure}
  \begin{center}
    \includegraphics[width=0.98\textwidth]{images/BSIS/cp2p}
    \caption{The painted boundary method with aggregate weight
    updates following the second iteration. In this iteration a
    painted vertex was selected, leading to the painting of its
    on-processor neighbors.}
    \label{5:fig:cp2-parallel}
  \end{center}
\end{figure}
The method finishes when the data structure is emptied of all
vertices. The product is now three disjoint sets: $C$ and $F$, as
usual, but also a set of painted vertices. The painted vertices are
the vertices that cannot be assigned to the $F$ or $C$ set without
communication with another processor. The information is provided to
CLJP-c, which handles the parallel portion of the algorithm.

The painted boundary approach is ideal given large numbers of interior
vertices, which can be guaranteed on most problems for the fine grid.
A side-effect of H1-based coarsening algorithms, however, is the
creation of denser graphs on coarse levels. One solution is to use the
painted boundary method on fine levels and then switch to CLJP-c
further along in the coarsening process. The issue is a smaller
concern for H1$^\prime$-based coarsening using BSIS since these
methods produce small operator complexities that do not grow as the
size of the problem is increased. In all cases, however, the number of
processor boundary vertices relative to the number of interior
vertices increases as the number of unknowns per processor decreases
(e.g., on coarse levels). A few techniques may be applicable in this
situation. The easiest solution is to simply use CLJP-c, or some other
coarsening algorithm, on the coarser levels where few vertices are
found. Although BSIS does not decrease in cost at this point, the
total cost on the levels when BSIS is not used is low due to low
complexity of coarse grids. A second approach is the application of a
dynamic load balancing algorithm to increase the number of vertices on
a processor (and, thus, decrease communication costs). If the number
of vertices per processor per level is maintained at a high enough
level, BSIS is still valuable on coarse grids. A third option is to
replicate the operator matrix on processors, which leads to processors
doing more of the same work, but by avoiding communication. The second
and third ideas are similar in nature and both involve using dynamic
load balancing
techniques~\cite{DevineBomanKarypisPP04,CybDLB,deCoughnyLB,SchloegelLB,ZoltanParHyp06ipdps,ZoltanParHypRepart07}.

\bibliographystyle{elsart-num}
\bibliography{paper}
